Effects of bariatric surger and weight loss on resting-state functional connectivity

### 2020-12-01 draft_rsfmri.Rmd
update novelty statement

<!-- obesity and interventions -->
Over the past decades, the global burden of obesity has increasingly been confronted as a matter of urgency by research.
_something about comorbidities_
Besides metabolic dysfunctions, obesity has been consistently associated with reduced gray matter volume, less intact white matter structure and reduced cognitive function [@Beyer_2019; @Zhang_2018]. 
Luckily, dietary interventions such as caloric restriction and related metabolic improvements, and weight loss have been suggested to improve neuronal plasticity and cognitive performance [@Witte_2009; @Zhang_2018; @Veronese_2017].
In some cases behavioral intervention fail to yield the desired success regarding weight loss. There are non-behavioral alternatives at hand.
<!-- bars most invasive (brain metabolism), last resort -->
Bariatric surgery is one promising option to combat morbid obesity, as it rapidly improves weight status, metabolic dysfunctions, and co-morbidities, such as diabetes. Roux-en-Y gastric bypass (RYGB) is the most common procedure as it not only restricts the food intake by reducing stomach volume but also affects the absorption of macronutrients and the hormonal profile [@]. <!-- ref for "common", ref for procedure -->
Other widely used procedures are vertical sleeve gastrectomy (VSG) and gastric banding (GB) which only restrict food consumption. <!-- mehr zu Veränderung in Gasto-intestinaler Physiologie -->

@@

<!-- Default mode network -->

Weight regulation and energy expenditure is dependent on a complex interplay of external and internal sensory processing, as well as molecular homeostasis regulation. Obesity may be cause or effect of a decompensation of the psychophysiological system that ensures homeostasis.
The default mode network is commonly associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> It is mainly composed of the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (MPFC) and inferior and lateral parietal cortex (IPC, LPC; approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (VACC), superior frontal cortex (SFC), and inferior temporal cortex (ITC) show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left VACC, middle temporal gyrus (MTG), and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- difference lean obese --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2016]. Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
A longitudinal study that confirmed the stronger connectivity (of PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG). However, default mode connectivity in these regions seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable 1 year after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
This is consistent with other findings of reduction of functional connectivity of these and other regions of the default mode network, such as the VMPFC [@Li_2018] <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery.

@@

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing patients that only differs in regard to the intervention, i.e. undergoing bariatric surgery, over an extended time period.
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network (DMN).
Participants are measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008].

>>>>

<!-- obesity and interventions -->
Over the past decades, the global burden of obesity has increasingly been confronted as a matter of urgency by research.
_something about comorbidities_
Besides metabolic dysfunctions, obesity has been consistently associated with reduced gray matter volume, less intact white matter structure and reduced cognitive function [@Beyer_2019; @Zhang_2018]. 
Luckily, dietary interventions such as caloric restriction and related metabolic improvements, and weight loss have been suggested to improve neuronal plasticity and cognitive performance [@Witte_2009; @Zhang_2018; @Veronese_2017].
In some cases behavioral intervention fail to yield the desired success regarding weight loss. There are non-behavioral alternatives at hand.
<!-- bars most invasive (brain metabolism), last resort -->
Bariatric surgery is one promising option to combat morbid obesity, as it rapidly improves weight status, metabolic dysfunctions, and co-morbidities, such as diabetes [@]. Roux-en-Y gastric bypass (RYGB) is the most common procedure as it not only restricts the food intake by reducing stomach volume but also affects the absorption of macronutrients and the hormonal profile [@]. <!-- ref for "common", ref for procedure -->
Other widely used procedures are vertical sleeve gastrectomy (VSG) and gastric banding (GB) which only restrict food consumption. <!-- mehr zu Veränderung in Gasto-intestinaler Physiologie -->

@@

<!-- Default mode network --> 
Weight regulation and energy expenditure is dependent on a complex interplay of external and internal sensory processing, as well as molecular homeostasis regulation. Obesity may be cause or effect of a decompensation of the psychophysiological system that ensures homeostasis.
The default mode network is commonly associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> It is mainly composed of the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (MPFC) and inferior and lateral parietal cortex (IPC, LPC; approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (VACC), superior frontal cortex (SFC), and inferior temporal cortex (ITC) show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left VACC, middle temporal gyrus (MTG), and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- difference lean obese --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2016]. Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
A longitudinal study that confirmed the stronger connectivity (of PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG). However, default mode connectivity in these regions seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable 1 year after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
This is consistent with other findings of reduction of functional connectivity of these and other regions of the default mode network, such as the VMPFC [@Li_2018] <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery.

@@

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing a cohort of patients to a cohort with similar characteristics but unexposed to the intervention, i.e. bariatric surgery, over an extended time period as would be demanded to draw more substantiated inference [@George_2016].
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery by comparison with a contemporaneously measured waiting-control group. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network (DMN).
Participants are measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008].
